---
title: "Statistics in Action - case study 1"
author: "Sebastien Moeller"
output:
  html_document:
    fig_height: 3
    fig_width: 8
    number_sections: no
    toc: no
  pdf_document:
    fig_height: 3
    fig_width: 8
    includes:
      before_body: macros.tex
    keep_tex: yes
    number_sections: no
    toc: no
---

```{r}
if (!require("pacman")) install.packages("pacman")

pacman::p_load(equivalence, tidyverse, ggplot2, dplyr)
```

# Introduction

The dataset <ttt>dataMON810_2018.csv</ttt> consists of several measurements made during a subchronic toxicity study concerning the MON810 maize.

Biochemical parameters reflecting most physiological functions were measured two times (week 5 and 14), in particular through serum and urine chemistry, and hematology. Organ weights were measured at week 14.

The main objective of this study is to evaluate possible GMO effects on these parameters.

# Single comparison

#### __(Q1)__ 
We consider the variable "CALCIUM".
```{r}
data <- read.csv('dataMON810_2018.csv')
```

###### a. Test if the mean level of calcium for period 2 is the same for males and females.  </br>**Hint:** plot first the data and justify the test(s) to use.
```{r}
# Filter the data was want and remove NA from variable
calciumA <- data %>% filter(period == 2) %>% dplyr::select(CALCIUM, sex) %>% na.omit
# Calculate the sample means
calciumAMean <- aggregate(calciumA$CALCIUM  ~ calciumA$sex, FUN = 'mean')
colnames(calciumAMean) <- c('sex', 'CALCIUM')

# Plot the data
ggplot() + 
  geom_point(data = calciumA, aes(x = CALCIUM, y = sex, colour = sex, alpha = 0.5), show.legend = FALSE) +
  geom_point(data = calciumAMean, aes(x = CALCIUM, y = sex, color = sex, size = 4, alpha = 0.5), show.legend = FALSE)

# Let's take a look at the means
calciumAMean
```

Looking at the data we see that the `$CALCIUM$` means between male and female are close. Computing a p-value here is of interest to see if they are significantly different.

The hypotheses we are testing are as follow:
$$
H_0:\mu_{x} = \mu_{y} \space\text{ versus }\space H_1:\mu_{x} \neq\mu_{y}
$$

###### Assuming equal variances
We can use the function `t.test`. Let $x$ and $y$ denote $males$ and $females$ respectively.

Assuming first equal variances $(\sigma^2_x = \sigma^2_y)$. We select a 5% significance level and run the test.
```{r}
alpha <- 0.05
t.test((calciumA %>% filter(sex == 'M'))$CALCIUM, (calciumA %>% filter(sex == 'F'))$CALCIUM, conf.level = 1 - alpha, var.equal = TRUE)
```

Under the assumption that the two sexes have equal variances, we find a low `p-value < 2.2e-16` and thus __reject the null _hypothesis__ that $\mu_x = \mu_y$.

###### F-test for equal variances
As stated, the previous t-test assumed that the two populations have equal variances. An F-test is used to test if the variances of two populations are equal. We will perform a two tailed test to see if our previous assumption was valid or incorrect.

For the F-test, our hypotheses are as follow:
$H_0: \sigma_x = \sigma_y \space\text{versus}\space H_1:\sigma_x \neq \sigma_y$
```{r}
var.test((calciumA %>% filter(sex == 'M'))$CALCIUM, (calciumA %>% filter(sex == 'F'))$CALCIUM, conf.level = 1 - alpha, alternative = 'two.sided')
```

Our intuition was correct, the true ratio of variances $\frac{\sigma_x}{\sigma_y} \neq 1$. Thus we __reject the null hypothesis__.

###### Assuming different variances
Assuming equal variances between the two groups has been disputed.
```{r}
aggregate(calciumA$CALCIUM  ~ calciumA$sex, FUN = 'sd')
```

The sample data of size 97 for males and 90 for females have very different standard deviations. It is therefore important to run the t-test without the assumption of equal variances.
```{r}
t.test((calciumA %>% filter(sex == 'M'))$CALCIUM, (calciumA %>% filter(sex == 'F'))$CALCIUM, conf.level = 1 - alpha, var.equal = FALSE)
```

Our new t-test also __rejects the null hypothesis__ that the two groups share the same mean displaying a `p-value = 3.455e-15`.

###### Conclusion
We can safely conclude that the two sexes do indeed have different means by the means of a t-test without equal variances.

##### b. Test for the males if the mean level of calcium is the same for period 1 and period 2.
```{r}
# Filter the data was want and remove NA from variable
calciumB <- data %>% filter(sex == 'M') %>% dplyr::select(period, CALCIUM) %>% na.omit
calciumB$period <- as.factor(calciumB$period)

calciumBMean <- aggregate(calciumB$CALCIUM  ~ calciumB$period, FUN = 'mean')
colnames(calciumBMean) <- c('period', 'CALCIUM')

# Plot the data
ggplot() + 
  geom_point(data = calciumB, aes(x = CALCIUM, y = period, colour = period, alpha = 0.5), show.legend = FALSE) +
  geom_point(data = calciumBMean, aes(x = CALCIUM, y = period, color = period, size = 4, alpha = 0.5), show.legend = FALSE)

# Let's take a look at the means
calciumBMean
```

###### F-test for equal variances
As stated, the previous t-test assumed that the two populations have equal variances. An F-test is used to test if the variances of two populations are equal. We will perform a two tailed test to see if our previous assumption was valid or incorrect.

For the F-test, our hypotheses are as follow:
$H_0: \sigma_x = \sigma_y \space\text{versus}\space H_1:\sigma_x \neq \sigma_y$
```{r}
var.test((calciumB %>% filter(period == 1))$CALCIUM, (calciumB %>% filter(period == 2))$CALCIUM, conf.level = 1 - alpha, alternative = 'two.sided')
```

Our intuition was correct, the true ratio of variances $\frac{\sigma_x}{\sigma_y} \neq 1$. Thus we __reject the null hypothesis__.

###### Assuming different variances
Assuming equal variances between the two groups has been disputed.
```{r}
aggregate(calciumB$CALCIUM  ~ calciumB$period, FUN = 'sd')
```

```{r}
dim(filter(calciumB, period == 1))[1]
dim(filter(calciumB, period == 2))[1]
```


The sample data of size 98 for period 1 males, and 97 for period 2 males have very different standard deviations. It is therefore important to run the t-test without the assumption of equal variances.
```{r}
t.test((calciumB %>% filter(period == 1))$CALCIUM, (calciumB %>% filter(period == 2))$CALCIUM, conf.level = 1 - alpha, var.equal = FALSE)
```

Our new t-test also __rejects the null hypothesis__ that the two groups share the same mean displaying a `p-value = 3.113e-10`.

###### Conclusion
We can safely conclude that the male population between the two periods do indeed have different means.

##### c. Test for the males if the mean level of calcium for period 2 is the same for the control group and the MON810 group.
```{r}
# Filter the data was want and remove NA from variable
calciumC <- data %>% filter(sex == 'M') %>% filter(period == 2) %>% dplyr::select(regimen, CALCIUM) %>% filter(regimen %in% c('MON810', 'control')) %>% na.omit
calciumC$regimen <- as.factor(calciumC$regimen)

calciumCMean <- aggregate(calciumC$CALCIUM  ~ calciumC$regimen, FUN = 'mean')
colnames(calciumCMean) <- c('regimen', 'CALCIUM')

# Plot the data
ggplot() + 
  geom_point(data = calciumC, aes(x = CALCIUM, y = regimen, colour = regimen, alpha = 0.5), show.legend = FALSE) +
  geom_point(data = calciumCMean, aes(x = CALCIUM, y = regimen, color = regimen, size = 4, alpha = 0.5), show.legend = FALSE)

# Let's take a look at the means
calciumCMean
```

###### F-test for equal variances
As stated, the previous t-test assumed that the two populations have equal variances. An F-test is used to test if the variances of two populations are equal. We will perform a two tailed test to see if our previous assumption was valid or incorrect.

For the F-test, our hypotheses are as follow:
$H_0: \sigma_x = \sigma_y \space\text{versus}\space H_1:\sigma_x \neq \sigma_y$
```{r}
var.test((calciumC %>% filter(regimen == 'control'))$CALCIUM, (calciumC %>% filter(regimen == 'MON810'))$CALCIUM, conf.level = 1 - alpha, alternative = 'two.sided')
```

Our F-test returns a `p-value = 0.03762` and thus __rejects the nullhypothesis__ that $\frac{\sigma_x}{\sigma_y} = 1$ at a 5% two tailed confidence level.

###### Assuming different variances
Assuming equal variances between the two groups has been disputed.
```{r}
aggregate(calciumC$CALCIUM  ~ calciumC$regimen, FUN = 'sd')
```

```{r}
dim(filter(calciumC, regimen == 'control'))[1]
dim(filter(calciumC, regimen == 'MON810'))[1]
```


Out control set consists of 20 observations and the MON810 consists of 19 observations and have different standard deviations. It is therefore important to run the t-test without the assumption of equal variances.
```{r}
t.test((calciumC %>% filter(regimen == 'control'))$CALCIUM, (calciumC %>% filter(regimen == 'MON810'))$CALCIUM, conf.level = 1 - alpha, var.equal = FALSE)
```

Our new t-test  __rejects the alternate hypothesis__ that the two groups do not share the same mean displaying a very significant `p-value = 0.4466`.

###### Conclusion
We conclude that the two groups do not have significantly different means.



------------


##### d. What is the probability to detect a difference of 1 sd (one standard deviation) 


*i)* with only 10 animals per group?
As we only have two groups, namely: `control` and `MON810` with only 10 animals per group we have `df = 10*2-2`

*ii)* with 20 animals? 
As we only have two groups, namely: `control` and `MON810` with only 20 animals per group we have `df = 20*2-2`

*iii)* How can we ensure to detect such difference with a probability of 80%?

```{r}

```


--------------  

##### e. Test for the males if the mean levels of calcium for period 2 of the `control` and `MON810` groups* are equivalent. The equivalence limits will be defined using the 6 reference groups as 
```{r}
# Filter the relevant data
calciumE <- data %>% filter(sex == 'M', period == 2, regimen != 'control', regimen != 'MON810') %>% dplyr::select(regimen, CALCIUM) %>% na.omit
calciumE$regimen <- as.factor(calciumE$regimen)

# Calculate the mean of each regiem
calciumEMean <- aggregate(calciumE$CALCIUM  ~ calciumE$regimen, FUN = 'mean')
colnames(calciumEMean) <- c('regimen', 'CALCIUM')
calciumEMean
```

*i)* one standard deviation of the 6 reference means,
```{r}
# The size of our confidence interval is given by a deviation from the mean in either direction of:
sd(calciumEMean$CALCIUM)

# The probability of observing a deviation of this size
pt(q = -sd(calciumEMean$CALCIUM), df = dim(calciumE)[1]-6)*2*100

# Calculate the p-value
#tost((calciumC %>% filter(regimen == 'control'))[,2], (calciumC %>% filter(regimen == 'MON810'))[,2], alpha = 0.05, epsilon = sd(calciumEMean$CALCIUM))
```

Our degrees of freedom is given by the number of observations minus the number of separate sets we calculated the means from: `df = 58 - 6 = 52`
The result shows that we have a probability of `42.30%` of observing one standard deviation of the 6 reference means given all observations.

*ii)* two standard deviations of the 6 reference means.
```{r}
# The probability of observing a deviation of twice this size
pt(q = -2*sd(calciumEMean$CALCIUM), df = dim(calciumE)[1]-6)*2*100
```

Similarly, the probability of observing two standard deviations from the mean given our data is `11.75%` with `df = 52`
   
##### f. Summarize and comment these results.
```{r}

```















#### __(Q2)__ Do the same analysis with the variable "DIRBILI" (direct bilirubin)

We consider the variable "DIRBILI".
###### a. Test if the mean level of DIRBILI for period 2 is the same for males and females.
```{r}
# Filter the data was want and remove NA from variable
dirbiliA <- data %>% filter(period == 2) %>% dplyr::select(DIRBILI, sex) %>% na.omit

# Calculate the sample means
dirbiliAMean <- aggregate(dirbiliA$DIRBILI  ~ dirbiliA$sex, FUN = 'mean')
colnames(dirbiliAMean) <- c('sex', 'DIRBILI')

# Plot the data
ggplot() + 
  geom_point(data = dirbiliA, aes(x = DIRBILI, y = sex, colour = sex, alpha = 0.5), show.legend = FALSE) +
  geom_point(data = dirbiliAMean, aes(x = DIRBILI, y = sex, color = sex, size = 4, alpha = 0.5), show.legend = FALSE)
```

This data seems to be discrete, having a value of 0.100 or 0.001 in period 2.
```{r}
unique(dirbiliA)
```

```{r}
# Let's take a look at the means by sex
dirbiliAMean
```

We will make use of the Chi-Square test to test for independence of `gender` for `DIRBILI`. If gender is independent then the two sexes share the same mean, otherwise they have distinct means.

For the Chi-square test to function in R we need to transform the data into a contingency table following the form:
```{r}
table(dirbiliA)
```

We set `correction = FALSE` as correction for continuity adjusts the formula for Pearson's chi-squared test. It subtracts 0.5 from the difference between each observed value and its expected value in a 2 × 2 contingency table. This reduces the chi-squared value obtained and thus increases its p-value. This formula is chiefly used when at least one cell of the table has an expected count smaller than 5. As Yates' correction also tends to overcorrect, and we know that we have significantly more than 5 counts per group we choose not to perform the correction.

The Chi-square test is performed as follows:
$\chi^2=\sum_{i = 1}^N\frac{(O_i-E_i)^2}{E_i}$

The df is given by a function of the contingency table: `df = (nrow - 1)(ncol - 1) = (1)(1) = 1`

$H_0: \mu_x\neq\mu_y\space\text{versus}\space H_1:\mu_x=\mu_y$
```{r}
chisq.test(table(dirbiliA), correct = FALSE)
```

Our test returns `p-value = 0.5814` indicating that the two sexes DIRBILI values are not independent and actually share a common mean.

###### b. Test for the males if the mean level of DIRBILI is the same for period 1 and period 2.
```{r}
# Filter the data was want and remove NA from variable
dirbiliB <- data %>% dplyr::select(DIRBILI, period) %>% na.omit
dirbiliB$period <- as.factor(dirbiliB$period)

# Calculate means by period
dirbiliBMean <- aggregate(dirbiliB$DIRBILI ~ dirbiliB$period, FUN = 'mean')
colnames(dirbiliBMean) <- c('period', 'DIRBILI')

# Plot the data
ggplot() +
  geom_point(data = dirbiliB, aes(x = DIRBILI, y = period, colour = period, alpha = 0.5), show.legend = FALSE) +
  geom_point(data = dirbiliBMean, aes(x = DIRBILI, y = period, color = period, size = 4, alpha = 0.5), show.legend = FALSE)
```

We immediately suspect that the two periods do not have independent means. Let us verify our suspicion with a Chi-square test with the same set-up as question a.
```{r}
table(dirbiliB)
```

This Chi-square test does not need correction at the Yates' correction is only available on 2x2 contingency tables. For our test the `df = (nrow - 1)(ncol - 1) = (2)(1) = 2` derived from the contingency table.
```{r}
chisq.test(table(dirbiliB))
```

We calculate `p-value = 2.2e-16`, and thus reject the null hypothesis. The two periods have distinct means.

###### c. Test for the males if the mean level of DIRBILI for period 2 is the same for the control group and the MON810 group.
```{r}
# Filter the data
dirbiliC <- data %>% filter(sex == 'M', period == 2) %>% dplyr::select(regimen, DIRBILI) %>% filter(regimen %in% c('MON810', 'control')) %>% na.omit

# Calculate the means per regimen
dirbiliCMean <- aggregate(dirbiliC$DIRBILI ~ dirbiliC$regimen, FUN = 'mean')
colnames(dirbiliCMean) <- c('regimen', 'DIRBILI')

# Plot the data
ggplot() + 
  geom_point(data = dirbiliC, aes(x = DIRBILI, y = regimen, colour = regimen, alpha = 0.5), show.legend = FALSE) +
  geom_point(data = dirbiliCMean, aes(x = DIRBILI, y = regimen, color = regimen, size = 4, alpha = 0.5), show.legend = FALSE)
```

The plot seems to imply equal means between the `control` and `MON810` regimens.
```{r}
table(dirbiliC)[1:2,]
```

The contingency table shows that the two regiems have exactly the same frequency of each value. We expect a `p-value = 1`. As these are random variables, the true p-value is likely to be very close to 1
```{r}
chisq.test(table(dirbiliC)[1:2,])
```

With a `p-value = 1` we reject the alternative hypothesis. The two regimens `contol` and `MON810` share the same mean.

###### d. What is the probability to detect a difference of 1 sd (one standard deviation)

*i)* with only 10 animals per group? 

*ii)* with 20 animal? 

*iii)* How can we ensure to detect such difference with a probability of 80%?

###### e. Test for the males if the mean levels of calcium for period 2 of the control group and the MON810 are equivalent. The equivalence limits will be defined using the 6 reference groups as
```{r}
# Filter the data
dirbiliE <- data %>% filter(sex == 'M', period == 2, regimen != 'control', regimen != 'MON810') %>% dplyr::select(regimen, DIRBILI) %>% na.omit
dirbiliE$regimen <- as.factor(dirbiliD$regimen)

# Calculate the mean of each regiem
dirbiliEMean <- aggregate(dirbiliE$DIRBILI  ~ dirbiliE$regimen, FUN = 'mean')
colnames(dirbiliEMean) <- c('regimen', 'DIRBILI')
dirbiliEMean
```

*i)* one standard deviation of the 6 reference means
```{r}
# The size of our confidence interval is given by a deviation from the mean in either direction of:
sd(dirbiliEMean$DIRBILI)

# The probability of observing a deviation of this size
pt(q = -sd(dirbiliEMean$DIRBILI), df = dim(dirbiliE)[1]-dim(dirbiliEMean)[1])*2*100
```

We find that the probability of observing one standard deviation is 99.18%.

*ii)* two standard deviations of the 6 reference means.
```{r}
# The probability of observing a deviation of twice this size
pt(q = -2*sd(dirbiliEMean$DIRBILI), df = dim(dirbiliE)[1]-dim(dirbiliEMean)[1])*2*100
```

We find that the probability of observing two standard deviations is 98.35%.

###### f. Summarize and comment these results.









# Multiple comparisons
#### __(Q1)__
Redo the three tests of the previous section (questions a., b. and c.) for now comparing  the means of all the quantitative variables (see the annex 5 of the ANSES report <ttt>BIOT2009sa0285Ra.pdf</ttt> to know the type of each variable).

###### a. mean level for period 2 is the same for males and females. 
```{r}
questionA <- function(variable){
  # Filter the data was want and remove NA from variable
  dataNew <- data %>% filter(period == 2) %>% dplyr::select(sex, variable) %>% na.omit
  
  # F-test
  pVal <- var.test((dataNew %>% filter(sex == 'M'))[,2], (dataNew %>% filter(sex == 'F'))[,2], conf.level = 1 - 0.05, alternative = 'two.sided')$p.val
  
  # Check F-test result
  varEqual <- pVal > 0.05
  
  # Perform the t-test
  t.test((dataNew %>% filter(sex == 'M'))[,2], (dataNew %>% filter(sex == 'F'))[,2], conf.level = 1 - 0.05, var.equal = varEqual)$p.val
}
```

```{r}
questionA('CALCIUM')
```

###### b. test for the males if the mean level is the same for period 1 and period 2.
```{r}
questionB <- function(variable){
  # Filter the variable for males in period 1 and 2
  dataNew <- data %>% filter(sex == 'M') %>% dplyr::select(period, variable) %>% na.omit
  
  # Calculate F-test's p-value
  pVal <- var.test((dataNew %>% filter(period == 1))[,2], (dataNew %>% filter(period == 2))[,2], conf.level = 1 - 0.05, alternative = 'two.sided')$p.value
  
  # Check if the two periods have equal variances
  varEqual <- pVal > 0.05
  
  # Perform the t-test
  t.test((dataNew %>% filter(period == 1))[,2], (dataNew %>% filter(period == 2))[,2], conf.level = 1 - 0.05, var.equal = varEqual)$p.value
}
```

```{r}
questionB('CALCIUM')
```

###### c. test for the males if the mean level for period 2 is the same for the control group and the MON810 group.
```{r}
questionC <- function(variable){
  # Filter the data was want and remove NA from variable
  dataNew <- data %>% filter(sex == 'M') %>% filter(period == 2) %>% dplyr::select(regimen, variable) %>% filter(regimen %in% c('MON810', 'control')) %>% na.omit
  
  # Calculate F-test's p-value
  pVal <- var.test((dataNew %>% filter(regimen == 'control'))[,2], (dataNew %>% filter(regimen == 'MON810'))[,2], conf.level = 1 - 0.05, alternative = 'two.sided')$p.val
  
  # Check if the two groups have the same variance
  varEqual <- pVal > 0.05
  
  # Perform the t-test
  t.test((dataNew %>% filter(regimen == 'control'))[,2], (dataNew %>% filter(regimen == 'MON810'))[,2], conf.level = 1 - 0.05, var.equal = varEqual)$p.val
  
}
questionC('CALCIUM')
```

###### Merg data
Store the results (i.e. all the p-values) in a dataframe with one variable per row and four columns (name of the variable + three p-values). 

Non-quantitative variables: BILI, PROT, GLUCOSE, BLOOD, KETONE, UROBILI, GAMMAGT, TESTIS
```{r}
variable <- data  %>% dplyr::select(-GLUCOSE, -BLOOD, -BILI, -PROT ,-KETONE, -UROBILI, -GAMMAGT, -TESTIS, -id, -X, -regimen, -sex, -period) %>% colnames()

results <- data.frame()
for (i in 1:59){
  print(cbind(i, variable[i], questionC(variable[i])))
  #results <- rbind(results, cbind(variable[i], questionA(variable[i])))
}
results
```


```{r}
questionA(variable[59])

questionB(variable[20])
questionB(variable[21])
questionB(variable[52])
questionB(variable[53])
questionB(variable[54])
questionB(variable[55])
questionB(variable[56])
questionB(variable[57])
questionB(variable[58])
questionB(variable[59])

questionC(variable[59])
```

#### __(Q2)__
For each of the three tests, adjust the p-values using the Bonferroni  and the Benjamini-Hochberg corrections. How can we interpret these results?




























