---
title: "Statistics in Action - case study 1"
author: "Sebastien Moeller"
output:
  html_document:
    fig_height: 3
    fig_width: 8
    number_sections: no
    toc: no
  pdf_document:
    fig_height: 3
    fig_width: 8
    includes:
      before_body: macros.tex
    keep_tex: yes
    number_sections: no
    toc: no
---


# Introduction

The dataset <ttt>dataMON810_2018.csv</ttt> consists of several measurements made during a subchronic toxicity study concerning the MON810 maize.

Biochemical parameters reflecting most physiological functions were measured two times (week 5 and 14), in particular through serum and urine chemistry, and hematology. Organ weights were measured at week 14.

The main objective of this study is to evaluate possible GMO effects on these parameters.

# Single comparison

#### __(Q1)__ 
We consider the variable "CALCIUM".
```{r}
data <- read.csv('dataMON810_2018.csv')
```

###### a. Test if the mean level of calcium for period 2 is the same for males and females.  </br>**Hint:** plot first the data and justify the test(s) to use.
```{r}

```

###### b. Test for the males if the mean level of calcium is the same for period 1 and period ### 2.
```{r}

```

###### c. Test for the males if the mean level of calcium for period 2 is the same for the control group and the MON810 group.
```{r}

```

###### d. What is the probability to detect a difference of 1 sd (one standard deviation) with only 10 animals per group? with 20 animal? How can we ensure to detect such difference with a probability of 80%?
```{r}

```
  
###### e. Test for the males if the mean levels of calcium for period 2 of the control group and the MON810 are equivalent. The equivalence limits will be defined using the 6 reference groups as *i)* one standard deviation of the 6 reference means, *ii)* two standard deviations of the 6 reference means.
```{r}

```
   
###### f. Summarize and comment these results.
```{r}

```

#### __(Q2)__ Do the same analysis with the variable "DIRBILI" (direct bilirubin)




# Multiple comparisons
#### __(Q1)__
Redo the three tests of the previous section (questions a., b. and c.) for now comparing  the means of all the quantitative variables (see the annex 5 of the ANSES report <ttt>BIOT2009sa0285Ra.pdf</ttt> to know the type of each variable). Store the results (i.e. all the p-values) in a dataframe with one variable per row and four columns (name of the variable + three p-values). 

#### __(Q2)__
For each of the three tests, adjust the p-values using the Bonferroni  and the Benjamini-Hochberg corrections. How can we interpret these results?




























